\cleartorecto % A memoir-class command for moving the acknowledgments to a recto page, not verso.
\chapter{Acknowledgments} % For the heading on the page, also registers in the table of contents
\thispagestyle{plain} %This page should have numbers.

\epigraph{The Butcher would gladly have talked till next day,\\
But he felt that the Lesson must end,\\
And he wept with delight in attempting to say\\
He considered the Beaver his friend.}{Lewis Carroll, \textit{The Hunting of the Snark}}

As Isaac Newton famously did not say, \emph{``We are but dwarfs standing on the shoulders of somewhat bigger dwarfs.''} Looking back at these past five years, I cannot help but let this non-quote resonate in my heart: and indeed, I may not have seen much further; indeed, I may not really have stood that much -- but for whatever I did, I feel so grateful, and to so many. 

Of course, it is common practice to thank one's advisor, in my case Rocco Servedio. \emph{Common practice be damned:} I thank Rocco wholeheartedly, as the best advisor one could hope for and fathom, as the finest human being one could possibly envision. During these five years, he never made me feel dumb -- I was; he never complained about my bothering him -- I did; he was always thoughtful, smiling, knowledgeable, insightful; always available. I cannot possibly thank him enough -- here's only a poor attempt. Thank you, Rocco.

This set the bar quite high; yet, I was lucky enough to interact and be mentored by people above this bar, and way beyond. I specifically aim these thanks at Dana Ron, Ronitt Rubinfeld, and Madhu Sudan; who not only taught me so much, but also left me convinced that every single person in our field was wonderful, approachable, and intrinsically awesome. I hope I can live to their standard. I also am incredibly grateful to all my coauthors and collaborators: Jayadev Acharya, Tu\u{g}kan Batu, Eric Blais, Omri Ben-Eliezer, Anindya De, Ilias Diakonikolas, Themis Gouleakis, Elena Grigorescu, Talya Eden, Tom Gur, Venkat Guruswami, Daniel Kane, Gautam Kamath, Akash Kumar, Amit Levi, Reut Levi, Raghu Meka, Igor Carboni Oliveira, Alistair Stewart, Li-Yang Tan, and Karl Wimmer. I cannot promise I've learned from all of you -- I \emph{can} promise I tried, and wish I did.

I must also thank everybody that made my graduate life at Columbia so enjoyable and fulfilling, or, more pragmatically, even possible: Alex Andoni, Xi Chen, Fernando Krell, Igor Carboni Oliveira, Tal Malkin, Dimitris Paparas, Jessica Rosa, Erik Waingarten and -- of course! -- the coffee machine on the 5\textsuperscript{th} floor. Well, you know -- thank you. (Incidentally, Jessica, I sure hope I didn't drive you insane.) I must also thank the whole department at MIT; Gautam, again, and foremost; Ilya, Jerry, and everyone in CSAIL. You welcomed me, and all I had to offer was bad puns. Which brings me to my flatmates and friends in New York City and beyond. Thank you R\'emi, Joschi, Yumi, Laurent, Fay, and Juba; thank you, Vicky, and Narges. Thank you, Alaa. You put up with me for all these years, through all these jokes I'm pretty sure were not all funny: I don't know why, but I sure am grateful. Also, please, get a better sense of humor.

To my family -- Mom, Dad, Thomas, C\'{e}cile, Marion, Nicolas, Quentin, Margot, Sacha, and now (!) Lola. I will not thank you, the point is kind of moot -- you know what you are to me, what you have been all this time; and how could I possibly ever acknowledge so much? (Sorry, there is no joke here. I'm damn serious.)
Finally, to Nandini. For everything, and more.



% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \subsection{Acknowledgments (from distr-cc)}
% We thank Oded Goldreich and Rocco Servedio for insightful conversations and for technical and conceptual suggestions regarding the contents of this work and its presentation.
%
% \subsection{Acknowledgments (from cond)}
% We are sincerely grateful to the anonymous referees for their close reading of this paper
% and for their many helpful suggestions, which significantly improved the exposition of the
% final version.
