\chapter{Testing Classes of Distributions: Upper Bounds from Swiss Army Knives}\label{chap:unified:ub}

\epigraph{``Should we meet with a Jubjub, that desperate bird,\\
We shall need all our strength for the job!''}{Lewis Carroll, \textit{The Hunting of the Snark}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcounter{IRL}
\renewcommand{\theIRL}{\textsf{Succinctness}}
\newcommand*{\inlineref}[1]{\refstepcounter{IRL}({\theIRL})\label{#1}}
\newcommand{\RightComment}[1]{\Comment{\parbox[t]{.30\linewidth}{\small#1}}} % For algorithms
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this chapter, we focus on obtaining algorithmic upper bounds in distribution testing. Our goal, however, departs from most of the previous literature, in that it is not to solve a specific testing problem by coming up with a bespoke\cmargin{Idiomatic? Tailor-made?}{} algorithm for that task. We take instead a different path, and set out to provide \emph{general} algorithms applicable across-the-board to a variety of problems \emph{at once}. Marginally more formally, here is the objective we seek to address:
\begin{problem}\label{problem:ub:general}
Design general-purpose testing algorithms, that when applied to a property \property would have (tight, or near-tight, or not absolutely laughable) sample complexity $q(\eps, \tau)$ as long as \property satisfies some ``structural assumption'' $\mathcal{S}_\tau$ parameterized by $\tau$.
\end{problem}
\noindent For instance, one could think of $\mathcal{S}_\tau$ as ``every $\D\in\property$ is close to some piecewise-constant distribution on $1/\tau$ intervals'' (as is e.g. the case for monotone distributions, with $\tau=\frac{1}{\poly(\log n, 1/\eps)}$ by~\cref{theorem:Birge:obl:decomp}); or, in another vein, ``all $\D\in\property$ have $\norm{\D}_{17/4} \leq \tau$'' (technically, one \emph{could} think of that one).  

We make significant progress in this direction by providing two unified frameworks for the question of testing various properties of probability distributions. First, we describe in~\cref{sec:shaperestrictions} a meta-algorithm to test membership in any distribution class, particularly well-suited to any class (including monotone, log-concave, $t$-modal, piecewise-polynomial, and Poisson Binomial distributions) which satisfies a ``shape constraint.'' (Broadly speaking, whenever any distribution in the class is well-approximated, in a strong $\lp[2]$-type sense, by a piecewise-constant distribution on relatively few pieces.)

In~\cref{sec:fourier}, we detail our second general technique, based on an entirely different type of structural assumption. Namely, this approach now leverages purported properties of the \emph{Fourier transform} of the distributions, and performs particularly well for those classes containing distributions with sparse Fourier transform -- such as, for instance, the classes of Poisson Binomial distributions and SIIRVs.

Our two frameworks yield near-sample-optimal and computationally efficient testers for a wide range of distribution families; for most of these, we provide the first non-trivial tester in the literature.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Shape Restrictions Knife}\label{sec:shaperestrictions}
\input{chapter1/sec-shaperestrictions}

\section{The Fourier Knife}\label{sec:fourier}
\input{chapter1/sec-fourier}
